Actuator line modeling of vertical-axis turbines

### 2016-08-27 reviews/reviewer1-response.md

We use the overal rotor drag coefficient nomenclature ($C_D$) since this was
adopted by Paraschivoiu (2001). As noted, this is commonly called the thrust
coefficient for horizontal-axis turbines. We have added a note to the manuscript
to clarify this.

We have not plotted the LES results, since we only used LES for one tip speed
ratio, and for one turbine---it was mainly employed to assess wake predictions.

>Page 11, Line 42: I'm not quite sure if what you have stated about Smagorinsky
model is correct. It certainly is very easy to implement and gives nice results
in some cases, but we are all aware of its deficiencies. We set a fix Cs value
which means we add numerical dissipation to the whole computatiponal domain, but
this is not really accurate when dealing with cases with strong directional
homogeneity and anisotropy. Why didn't you choose a dynamic Smagorinsky or even
dynamic Lagrangian model? Besides, what is the Cs value you have used?

We are not trying to claim that the Smagorinsky model is the most accurate or
"the standard" model, but that is has been used in many other ALM papers, and
therefore a good baseline for comparison. The wording of the text has been
adjusted to reflect the notion that we are interested in this comparison to
baseline for turbine ALM modeling, and are by no means implying that this SGS
model is the most accurate or appropriate for this case. Certainly testing
additional SGS models will be part of future work, as is mentioned in the
conclusions.

The $C_S$ value used was OpenFOAM's default value of approximately 0.17. This
detail has been added to the text.

### 2016-08-27 reviews/reviewer1-response.md

>Page 13, Fig 15: Following my last comment, figure to the right may very well
explain that the Smagorinsly model has been over dissipative thereby hampering
turbulent fluctuations which may have otherwise resulted in higher TKEs.

This is a valid point, and we have added a comment to the text accordingly. It
is important to note, however, that since we don't have near-wall boundary
layers, it may be that turbulence is only generated by larger-scale shear and
vortex breakdown, rather than by boundary layer separation or transition, which
would occur with a body-fitted mesh.

>Page 14, Line 27: Matching mean velocities is usually rather straightforward
(in non-presudospectral codes, sometimes even shutting down the whole SGS
modeling wouln'd make a difference as long as no solid walls are involved). But
higher order moments such as Reynolds stress tensor components are usually more
sensitive. Perhaps you can go more in depth with that part -- possibly compare
your results with other experiments or CFD simulations. Besides, please try and
overcome the relatively large discrepancies seen in figures 6 (high TSR), 9, 11.

We think it is unlikely that the discrepancies seen in predicted loading are
related to higher order statistics of the flow field, as simplier blade element
methods, e.g., momentum and potential flow vortex methods, also often produce
similar overpredictions at higher TSR. It should also be noted that neither the
static nor dynamic blade element loading is computed as a function of the local
flow unsteadiness or TKE, though this would be an interesting avenue for future
investigation, which has never been attempted to our knowledge.

### 2016-08-27 paper.tex
Slim down intro a little; resolves #31

For analyzing turbine arrays, it is desirable to retain a Navier--Stokes
description of the flow field---in contrast to, e.g., momentum or potential flow
vortex models---to capture the effects of nonlinear advection and turbulent
transport. However, rather than resolving the fine details in the blade boundary
layers, an actuator-type model for parameterizing the turbine loading may be
employed, which dramatically drives down computational expense. As shown in
\cite{Bachant2015-JoT}, the conventional uniform actuator disk is not a good
candidate for a cross-flow turbine wake generator, never mind the fact that it
does not typically compute performance predictions. There are then two actuator
methods that use blade element theory to compute blade loading and therefore
generate a wake that more closely resembles that of an actual turbine. These are
the actuator cylinder or swept-surface model (ASSM) and the actuator line model
(ALM). The ASSM solves for the average blade loading along its path and applies
this as a constant body force term in the momentum equation. The ALM takes a
similar approach but is an unsteady method, resolving the blade element
locations in time.

The ALM, originally developed by Sorensen and Shen \cite{Sorensen2002}, has
become popular for modeling axial-flow or horizontal-axis turbines, and has been
shown in blind tests to be competitive with blade-resolved
CFD~\cite{Krogstad2013, Pierella2014}. The ALM combined with large eddy
simulation has become the state-of-the-art for modeling entire wind
farms~\cite{Archer2013, Churchfield2012, Sorensen2015, Fleming2013,
Fleming2014}. Like other blade element techniques, the effectiveness of the ALM
for AFTs is in part due to the quasi-steady nature of the flow in the blade
reference frame, and the relatively rare occurrence of stall. Note that a
similar method can be used with the Navier--Stokes equations in
vorticity--velocity form~\cite{Scheurich2011b}, which may be more efficient when
vorticity is confined to a relatively small fraction of the domain, e.g., for a
standalone turbine with a uniform laminar inflow.

The ASSM and ALM were implemented to model a very low Reynolds number (based on
rotor diameter, $Re_D \sim 10^4$) 2-D cross-flow turbine experiment in a flume
using large eddy simulation (LES)~\cite{Shamsoddin2014}. Performance predictions
for this case were not reported, but the ALM was shown to be more effective at
postdicting the wake characteristics measured in the experiments by Brochier et
al.~\cite{Brochier1986}. In this study we validated the ALM with experimental
data taken at Reynolds number two orders of magnitude higher, reporting its
performance predictions, neither of which have been done before. Additionally,
assessed the use of the ALM in more computationally affordable unsteady RANS
simulations, which also has not yet been reported for VATs.

It is therefore proposed that an actuator line model may be the optimal
combination of high-fidelity flow modeling that includes performance
predictions, but with reduced computational expense. Here we have developed an
ALM for cross-flow turbines inside both RANS and LES simulations, implemented as
an extension for the OpenFOAM open-source CFD library. This model was validated
against experimental data for the high solidity ($c/R=0.28$) UNH Reference
Vertical-Axis Turbine (UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US
Department of Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow
turbine (at 1:6 scale), both of which are shown in Figure~\ref{fig:turbines}.
Validation datasets for both turbines were taken from
\cite{Bachant2016-RVAT-Re-dep} and \cite{Bachant2016-RM2-data}, respectively.
The turbines were modeled at $Re_D \sim 10^6$, at which experimentally measured
performance and near-wake characteristics were nearly Reynolds number
independent~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}.

>>>>

For analyzing turbine arrays, it is desirable to retain a Navier--Stokes
description of the flow field---in contrast to, e.g., momentum or potential flow
vortex models---to capture the effects of nonlinear advection and turbulent
transport. However, rather than resolving the fine details in the blade boundary
layers, an actuator-type model for parameterizing the turbine loading may be
employed, which dramatically drives down computational expense. As shown in
\cite{Bachant2015-JoT}, the conventional uniform actuator disk is not a good
candidate for a cross-flow turbine wake generator, never mind the fact that it
does not typically compute performance predictions.

Actuator line modeling (ALM), originally developed by Sorensen and Shen
\cite{Sorensen2002}, is an unsteady method that tracks blade element locations,
and has become popular for modeling axial-flow or horizontal-axis turbines, and
has been shown in blind tests to be competitive with blade-resolved
CFD~\cite{Krogstad2013, Pierella2014}. The ALM combined with large eddy
simulation has become the state-of-the-art for modeling entire wind
farms~\cite{Archer2013, Churchfield2012, Sorensen2015, Fleming2013,
Fleming2014}. Like other blade element techniques, the effectiveness of the ALM
for AFTs is in part due to the quasi-steady nature of the flow in the blade
reference frame, and the relatively rare occurrence of stall. Note that a
similar method can be used with the Navier--Stokes equations in
vorticity--velocity form~\cite{Scheurich2011b}, which may be more efficient when
vorticity is confined to a relatively small fraction of the domain, e.g., for a
standalone turbine with a uniform laminar inflow.

The ASSM and ALM were implemented to model a very low Reynolds number (based on
rotor diameter, $Re_D \sim 10^4$) 2-D cross-flow turbine experiment in a flume
using large eddy simulation (LES)~\cite{Shamsoddin2014}. Performance predictions
for this case were not reported, but the ALM was shown to be more effective at
postdicting the wake characteristics measured in the experiments by Brochier et
al.~\cite{Brochier1986}.

In this study we have developed an ALM for cross-flow turbines and embedded the
model inside LES and unsteady RANS simulations, the latter of which has not yet
been reported in the literature. This model was validated against experimental
data for the high solidity ($c/R=0.28$) UNH Reference Vertical-Axis Turbine
(UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US Department of
Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow turbine (at 1:6
scale), both of which are shown in Figure~\ref{fig:turbines}. Validation
datasets for both turbines were taken from \cite{Bachant2016-RVAT-Re-dep} and
\cite{Bachant2016-RM2-data}, respectively. The turbines were modeled at $Re_D
\sim 10^6$, at which experimentally measured performance and near-wake
characteristics were nearly Reynolds number
independent~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}, and two orders of
magnitude higher than previous ALM investigations for VATs.
